### 2025-04-21 rebuttal.tex

\documentclass[10pt,twocolumn,letterpaper]{article}
\usepackage[rebuttal]{cvpr}

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}

% Import additional packages in the preamble file, before hyperref
\input{preamble}

% If you comment hyperref and then uncomment it, you should delete
% egpaper.aux before re-running latex.  (Or just hit 'q' on the first latex
% run, let it finish, and you should be clear).
\definecolor{cvprblue}{rgb}{0.21,0.49,0.74}
\usepackage[pagebackref,breaklinks,colorlinks,allcolors=cvprblue]{hyperref}

% If you wish to avoid re-using figure, table, and equation numbers from
% the main paper, please uncomment the following and change the numbers
% appropriately.
%\setcounter{figure}{2}
%\setcounter{table}{1}
%\setcounter{equation}{2}

% If you wish to avoid re-using reference numbers from the main paper,
% please uncomment the following and change the counter value to the
% number of references you have in the main paper (here, 100).
%\makeatletter
%\apptocmd{\thebibliography}{\global\c@NAT@ctr 100\relax}{}{}
%\makeatother

%%%%%%%%% PAPER ID  - PLEASE UPDATE
\def\paperID{16006} % *** Enter the Paper ID here
\def\confName{CVPR}
\def\confYear{2025}

\input{sec/custom_commands}
\begin{document}

%%%%%%%%% TITLE - PLEASE UPDATE
\title{Towards certifiable outlier-robust point cloud registration in real-time using intersection graphs}  % **** Enter the paper title here

\maketitle
\thispagestyle{empty}
\appendix

We first want to thank the reviewers for their comments, feedback and valuable ideas.
We are happy to hear that reviewer 1 enjoyed reading our paper, and that reviewer 1 and 3 think the ideas in our paper may inspire the research community.

\textbf{Response to reviewer 1}. To address the overall confusion about the problem setting reviewer 1 seemingly had, we realize that we will need to elaborate more on our specific use case which is global localization for autonomous driving, using LiDAR-sensors. For this use-case, only rigid transformation is needed (we will change the title to mention rigid), and in combination with local feature descriptors, a few hundred points are usually enough. We agree here fully with reviewer 1 that different use-cases have different requirements on scalability and speed. We think that considering there is a high-frequency odometry, doing LiDAR-registration with 10Hz (100ms) is the bare-minimum that could still count as real-time in our use-case.
The assumption of pre-made correspondences indeed means that we cannot just plug in raw point clouds and is therefore a limitation. Instead, we need to use keypoints and match them with feature descriptors, this was done in the evaluation section.
We fully agree with reviewer 1 that knowing what influence epsilon has on the estimation error \cite{antonante2024-robust-perception} would be valuable, we will add an evaluation for this.

\textbf{Response to reviewer 2}.
Addressing the main concern of novelty of reviewer 2, we want to point out that while on the surface, our approach looks similar to \cite{zhang20233d} because it uses maximal cliques, our approach is mathematically rigorous by stating it as an optimization problem with an exactly defined objective function, namely the Truncated Least Squares (TLS). It offers in our opinion a great insight why listing maximal cliques works well in practice, instead of only empirically observing this fact. Neither was it previously described what compatibility criterion really means (namely pairwise intersection of the so-called T-sets),
nor what the connection is between the global minimas of the Truncated Least Squares (TLS) objective and the maximal cliques of the newly introduced intersection graph. Also, when talking about global optimality, we have shown that maximum cliques are not a sufficient condition for global optimality due to collections of SE3- T-sets lacking the Helly-property, a key insight about the combinatorics of the problem that was previously overlooked.
\textbf{Question 2}: The answer is unfortunately no, as we wrote at the end of section 5.1.2.\\
%\textit{Unfortunately, being only able to reject some and not all instances means our solver is currently neither globally optimal (being able to find the global minimum of every instance and outputting a valid lower bound on every instance), nor certifiable (outputting only a valid lower bound on every instance)}.\\
\textbf{Question 3}: GNC definitively undermines global optimality because it is a heuristic, yes. But we really mean GNC often \textit{finds} the global optimum, not that it \textit{proves} that the found solution is the global optimum.\\
\textbf{Question 4}: We indeed have a bad performance but only with the FPFH descriptor on 3DMatch, we will investigate the root cause of this. Since we have a good performance on 3DMatch using FCGF descriptor (83.92\%) and on KITTI even an accuracy comparable to state-of-the art with both FPFH and FCGF (95.5\% and 96.76\% respectively, top-three overall), this shows our algorithm is able to compete with state-of-the-art with both descriptors at least on KITTI. We suspect a wrong parameter in the config or a bug in one of the evaluation scripts for which we ran out of time for fixing. \textbf{Question 5.1}: We will report the runtime on the real datasets KITTI and 3DMatch, we indeed forgot to include them. \textbf{Question 5.2}: We will consider adding synthetic experiments of common other registration algorithms. The purpose of the evaluation section 5.1 is mainly for comparing the only available globally optimal solver STRIDE regarding registration accuracy. Also, the solver STRIDE is extremely slow and not scalable at all for more than 100 points, this is why we cannot use it in the experiments 5.2 and 5.3.
\textbf{Question 5.3}: We will attempt to do at least 10, maybe 50.\\
\textbf{Question 5.4}: Due to random subsampling, using less points may only make our algorithm look less accurate compared to the others, not more accurate. But we fully agree that the experiments should be done under the same conditions for comparability. We will attempt to reproduce all the numbers ourselves under the same conditions instead of reporting from \cite{zhang20233d}.
We will add qualitative results (i.e. example of registered point clouds) if space permits.

\textbf{Response to reviewer 3}. We indeed did not tune the parameters and still got top-three results on KITTI compared to SoTa. We are confident that we will find the issue with the FPFH descriptor on 3DMatch, and obtain overall more consistent evaluation results that we will also discuss more thoroughly. We will add the info how much time each submodule of the algorithm takes.
Regarding Figure 1 missing explanations, we indeed see now that identifiers like $\vc_i$, $\vt$ and $\epsilon$ are not present in the illustration, we apologize for the confusion. We do not see any problem with neither Eq. (10) nor Eq. (11), could you please elaborate ?

%%%%%%%%% REFERENCES
{
    \small
    \bibliographystyle{ieeenat_fullname}
    \bibliography{main}
}

\end{document}

### 2025-04-21 sec/X_suppl.tex
Added from MA

\begin{theorem}[\textbf{Knowing the ground-truth inliers solves the TLS-problem}]
	Given the ground-truth inlier set $\mathcal{I}^*$, the global optimum $(\MR^*, \vt^*)$ is found by solving the least-squares problem:
	\begin{equation}
		\begin{aligned}
			(\MR^*, \vt^*) = \argmin_{(\MR, \vt) \in \SEthree} \sum_{i \in \mathcal{I}^*} \normsq{\MR \vp_i - \vq_i + \vt}
		\end{aligned}
	\end{equation}
\end{theorem}

\begin{proof}
    \label{proof:gt-corrs-solves-tls}
	Since we know the ground-truth inlier set, we know which residuals are less or equal $\epssq$ and therefore not truncated. Since $\min(x, \epssq) = x$ for all $x \leq \epssq$, we can equivalently rewrite the TLS-objective \ref{eq:pcr-tls} as:
    \begin{equation}
	\begin{aligned}
		&\sum_{i=1}^{N} \min \left(\normsq{\MR \vp_i - \vq_i + \vt}, \epsilon^2 \right)\\
        &= \sum_{i \in \mathcal{I}^*} \normsq{\MR \vp_i - \vq_i  + \vt}  + \sum_{i \in \{1, \cdots, N\}  \setminus \mathcal{I}^*} \epsilon^2
	\end{aligned}
    \end{equation}
    Since the second sum is constant, it does not affect the minimizer of the least-squares term (the first sum). Therefore, minimizing only  the first, i.e. solving the least-squares problem solves the TLS-problem.
\end{proof}

@@

\begin{lemma}[\textbf{Least-squares solution with two points}]\label{lemma:ls-two-points}
	In the special case of two points, the least squares solution for point cloud registration is
	\begin{equation}
		\begin{aligned}
			\min_{(\MR, \vt) \in \SEthree} & \normsq{\MR \vp_i - \vq_i  + \vt} +  \normsq{\MR \vp_j - \vq_j  + \vt}\\
			&= \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2
		\end{aligned}
	\end{equation}
	Both residual terms are equal at the minimum.
\end{lemma}

\begin{proof}
	A well known result is that the optimal translation is related to the optimal rotation and the centroids of both point clouds (see e.g \cite{Kabsch-1976-Point-set-alignment, sorkine2017least}):
	\begin{equation}
		\begin{aligned}
			\vt &= \bar{\vq} - \MR \bar{\vp}\\
			\bar{\vp} &= \frac{\vp_i + \vp_j}{2}, \, \bar{\vq} = \frac{\vq_i + \vq_j}{2}
		\end{aligned}
	\end{equation}
	We insert the optimal translation into the objective and simplify:
	\begin{equation}\label{eq:insert-t-opt}
		\begin{aligned}
			&\normsq{\MR \vp_i - \vq_i  + (\bar{\vq} - \MR \bar{\vp})} +  \normsq{\MR \vp_j - \vq_j  + (\bar{\vq} - \MR \bar{\vp})}\\
			=&\normsq{\MR (\vp_i - \bar{\vp}) - (\vq_i -\bar{\vq})} + \normsq{\MR (\vp_j - \bar{\vp}) - (\vq_j -\bar{\vq})} \\
			=&\normsq{\MR \left(\vp_i - \frac{\vp_i + \vp_j}{2} \right) - \left(\vq_i - \frac{\vq_i + \vq_j}{2} \right)}\\
			+ &\normsq{\MR \left(\vp_j - \frac{\vp_i + \vp_j}{2} \right) - \left(\vq_j -\frac{\vq_i + \vq_j}{2} \right)} \\
			=&\normsq{ \frac{1}{2}(\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j))}\\
			+ &\normsq{ \frac{1}{2}(\MR (\vp_j - \vp_i ) - (\vq_j - \vq_i))}\\
			=& 2\normsq{ \frac{1}{2}(\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j))}\\
			=& \frac{1}{2}\normsq{\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j)}\\
		\end{aligned}
	\end{equation}
	We just showed that both residual terms are equal at the optimal translation.	
	What remains is to optimize over the rotation $\MR$. By lemma \ref{lemma:ls-rot}, the minimum over rotations is:
	\begin{equation}
		\begin{aligned}
			&\min_{(\MR, \vt) \in \SEthree} \normsq{\MR \vp_i - \vq_i  + \vt} +  \normsq{\MR \vp_j - \vq_j  + \vt}\\
			&=\min_{\MR \in \SOthree} \frac{1}{2}\normsq{\MR (\vp_i - \vp_j ) - (\vq_i - \vq_j)}\\
			&= \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2
		\end{aligned}
	\end{equation}
\end{proof}

\begin{theorem}[\textbf{Non-empty intersection of T-Sets is equivalent to compatible TIMs}]
	\begin{equation}
		\begin{aligned}
			&\mathcal{T}_i \cap \mathcal{T}_j \neq \emptyset\\
			\iff &\exists (\MR, \vt) \in \SEthree:\\
			&\norm{\MR \vp_i - \vq_i + \vt} \leq \epsilon\\
			\land &\norm{\MR \vp_j - \vq_j + \vt} \leq \epsilon\\
			\iff &| \norm{\va_{ij}} - \norm{\vb_{ij}} | \leq 2 \epsilon
		\end{aligned}
	\end{equation}
	where 
	\begin{equation}
		\label{eq:tims}
		\begin{aligned}
			\va_{ij} &= \vp_i - \vp_j\\
			\vb_{ij} &= \vq_i - \vq_j
		\end{aligned}
	\end{equation}
\end{theorem}

\begin{proof}
	\label{proof:non-empty-intersection-implies-tims-compat}
	Whether there exists a $(\MR, \vt) \in \SEthree$ such that $\norm{\MR \vp_i - \vq_i  + \vt} \leq \epsilon$ is equivalent to whether the minimum is $\leq \epsilon$, i.e. whether $\min_{(\MR, \vt) \in \SEthree} \norm{\MR \vp_i - \vq_i  + \vt} \leq \epsilon$ holds.
	By lemma \ref{lemma:ls-two-points} by minimizing the sum of two residuals, at the minimum they are both equal and therefore if their sum is $\leq 2\epssq$, they both are $\leq \epssq$. We conclude therefore
	\begin{equation}
		\begin{aligned}
			&\mathcal{T}_i \cap \mathcal{T}_j \neq \emptyset \iff\\
			&\min_{(\MR, \vt) \in \SEthree} \normsq{\MR \vp_i - \vq_i  + \vt} + \normsq{\MR \vp_j - \vq_j  + \vt} \leq 2\epssq\\
			& \iff \frac{1}{2} \left( \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} \right)^2 \leq 2\epssq\\
			& \iff | \norm{\vp_i - \vp_j} - \norm{\vq_i - \vq_j} | \leq 2\epsilon\\
			& \iff | \norm{\va_{ij}} - \norm{\vb_{ij}} | \leq 2 \epsilon
		\end{aligned}
	\end{equation}
\end{proof}

\begin{lemma}[\textbf{Consensus sets are cliques in the intersection graph}]
	Consensus sets are cliques in the intersection graph.
\end{lemma}

\begin{proof}
	Cliques in the intersection imply by definition pairwise non-empty intersection of T-Sets. Since pairwise non-empty intersection is a necessary condition for total non-empty intersection and therefore consensus sets, consensus sets are cliques in the intersection graph.
\end{proof}

\begin{lemma}[\textbf{Maximal clique enumeration and global optimality}]
	\label{lemma:max-clique-non-helly-app}
	Maximal clique enumeration finds the ground-truth consensus set and therefore the global minimum of an instance if all maximal cliques of this instance are maximal consensus sets.
\end{lemma}

\begin{proof}
	The key statement of the lemma is that if every maximal clique is a maximal consensus set, then subcliques of maximal cliques (which by definition are not maximal cliques) cannot be the ground truth inlier set. Consequently, they need not be enumerated.

	The global optimum lies in the union of all T-sets, i.e. $\bigcup_{i \in \RangeOneToN} \mathcal{T}_i$. Every vertex of a graph is contained in at least one maximal clique of that graph. Therefore, the global optimum is contained in at least one union $\bigcup_{i \in \vc} \mathcal{T}_i$ of a maximal clique $\vc$.

	The intersection of the T-sets of any subclique contains the intersection of the T-sets of a maximal clique containing that subclique. Therefore, if the consensus set is maximal, the best solution (minimizing the TLS objective) that can be found among subcliques is no better than this maximal clique. If this holds for every maximal clique of an instance, then the ground-truth inlier set is not a subclique of any maximal clique, and therefore must itself be a maximal clique. Thus, enumerating maximal cliques will find the ground-truth consensus set if all maximal cliques of this instance are maximal consensus sets. v
\end{proof}

\label{proof:wls-interval-analysis-rotation}

\begin{lemma}[\textbf{Consensus set size and TLS objective }]
	Given an upper-bound $\text{UB}$ on the TLS objective, the ground-truth consensus set must have at least $d_{m}$:
    \begin{equation}
        \begin{aligned}
            d_m = \lceil N - \text{UB} / \epssq \rceil
        \end{aligned}
    \end{equation}
\end{lemma}

\begin{proof}
    By the definition of the TLS objective, 
    $r_{in} + (N - d_m) \epsilon^2 \leq \text{UB}$ must hold, where $r_{in}$ is the (unknown) sum of the inlier residuals. Rearranging yields $d_m \geq r_{in}  + N - \text{UB} / \epssq$.
    Since all the summands in the TLS objective are non-negative, $r_{in} = 0$ is a valid lower bound. Since the number of elements in the consensus set $d_m$ must be an integer, we conclude that $d_m = \lceil N - \text{UB} / \epssq \rceil$.
\end{proof}

>>>>

@@

\section{Interval analysis of residuals under translation}
\label{proof:tls-relax-minmax-r}

\begin{theorem}
	The residual term $r_i = ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{t}||_2^2$
	has the minimum $r_{min}^i = \max(0, ||\mathbf{d}||_2 - n_r)^2$ and the maximum $r_{max}^i = (||\mathbf{d}||_2 + n_r)^2$
	for any translation $\mathbf{t} \in \Rthree, ||\mathbf{c} - \mathbf{t}|| \leq n_r$ where $\mathbf{d} = \mathbf{a}_i - \mathbf{b}_i + \mathbf{c}$.
\end{theorem}

\begin{proof}
	With $\mathbf{t}= \mathbf{c} - \Delta \mathbf{t} \iff \Delta \mathbf{t} = \mathbf{c} - \mathbf{t}$:
	\begin{equation}
		\begin{aligned}
			r_i &= ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{t}||_2^2
			= ||\mathbf{a}_i - \mathbf{b}_i + \mathbf{c} - \Delta\mathbf{t}||_2^2
			= ||\mathbf{d} - \Delta\mathbf{t}||_2^2\\ 
			&= (\mathbf{d} - \Delta\mathbf{t})^\transposed(\mathbf{d} - \Delta\mathbf{t})=
			\mathbf{d}^\transposed\mathbf{d} - 2 \mathbf{d}^\transposed \Delta\mathbf{t} + \Delta\mathbf{t}^\transposed \Delta\mathbf{t}\\
			&= ||\mathbf{d}||_2^2 - 2 \, ||\mathbf{d}|| \, ||\Delta\mathbf{t}||  \cos \theta + ||\Delta\mathbf{t}||_2^2\\
		\end{aligned}
	\end{equation}

	\begin{equation}
		\begin{aligned}
			r_{min}^i = \min_{\Delta \mathbf{t}} r_i &= 
			||\mathbf{d}||_2^2 + 
			\min_{\theta \in [-\pi, \pi], ||\Delta\mathbf{t}|| \in [0, n_r]} \left( -2 \, ||\mathbf{d}||  \, ||\Delta\mathbf{t}|| \cos \theta + ||\Delta\mathbf{t}||_2^2 \right) \\
			&= ||\mathbf{d}||_2^2 + \min_{||\Delta\mathbf{t}|| \in [0, n_r]} \left( -2 \, ||\mathbf{d}||  \, ||\Delta\mathbf{t}|| + ||\Delta\mathbf{t}||_2^2 \right) \\
			&= \min_{||\Delta\mathbf{t}|| \in [0, n_r]}  (||\mathbf{d}||_2 - ||\Delta\mathbf{t}||_2)^2\\
			&=\max(0, ||\mathbf{d}||_2 - n_r)^2\\
			r_{min}^i = \max_{\Delta \mathbf{t}} r_i &= 
			||\mathbf{d}||_2^2 + 
			\max_{\theta \in [-\pi, \pi], ||\Delta\mathbf{t}|| \in [0, n_r]} \left( -2 \, ||\mathbf{d}||  \, ||\Delta\mathbf{t}|| \cos \theta + ||\Delta\mathbf{t}||_2^2 \right) \\
			&= ||\mathbf{d}||_2^2 + 2 \, ||\mathbf{d}||\, n_r + n_r^2 = (||\mathbf{d}||_2 + n_r)^2
		\end{aligned}
	\end{equation}

\section{Interval analysis of residuals under rotation}
\label{proof:wls-interval-analysis-rotation}

\begin{theorem}

	The residual term $r_i = ||\mathbf{a}_i - \mathbf{R}\mathbf{b}_i ||_2^2$
	has the minimum $r_{min}^i$ and the maximum $r_{max}^i$
	over all 3D-rotations $\mathbf{R} \in \SOthree$ that satisfy $d_{\angle}(\mathbf{R}, \mathbf{R_c}) \leq n_r$ where $\mathbf{R_c} \in \SOthree$ is the ball center:
	\begin{equation}
		\begin{aligned}
			r_{min}^i &= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\max(\theta - n_r, 0))\\
			r_{max}^i &= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\min(\theta + n_r, \pi))
		\end{aligned}
	\end{equation}
	with 
	\begin{equation}
		\begin{aligned}
			\theta &= \arccos \left( \frac{(\mathbf{R}_c^\transposed \, \mathbf{a}_i )^\transposed \mathbf{b}_i}
			{||\mathbf{a}_i|| \, || \mathbf{b}_i||} \right)
		\end{aligned}
	\end{equation}
	being the angle between $\mathbf{R}_c^\transposed \, \mathbf{a}_i $ and $\mathbf{b}_i$.

\end{theorem}

\begin{proof}
	With $\mathbf{R}  = \mathbf{R}_c \Delta\mathbf{R} \iff \mathbf{R}_c^\transposed\mathbf{R} = \Delta \mathbf{R}$:
	\begin{equation}
		\begin{aligned}
			r_i &= ||\mathbf{a}_i - \mathbf{R}\mathbf{b}_i ||_2^2 \\
			&= || \mathbf{R}_c^\transposed \left(\mathbf{a}_i - \mathbf{R}\mathbf{b}_i\right) ||_2^2\\
			&= || \mathbf{R}_c^\transposed \mathbf{a}_i - \mathbf{R}_c^\transposed \mathbf{R}\mathbf{b}_i ||_2^2\\
			&= \left(\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta \mathbf{R}\mathbf{b}_i \right)^\transposed
			\left(\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta \mathbf{R}\mathbf{b}_i \right)\\ 
			&= \left(\mathbf{R}_c^\transposed \mathbf{a}_i \right)^\transposed \left(\mathbf{R}_c^\transposed \mathbf{a}_i \right) - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i + (\Delta \mathbf{R}\mathbf{b}_i )^\transposed (\Delta \mathbf{R}\mathbf{b}_i )\\
			&= \mathbf{a}_i^\transposed \mathbf{R}_c \mathbf{R}_c^\transposed \mathbf{a}_i - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i + \mathbf{b}_i^\transposed  \Delta \mathbf{R}^\transposed  \Delta \mathbf{R}\mathbf{b}_i\\
			&= \mathbf{a}_i^\transposed \mathbf{a}_i - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i + \mathbf{b}_i^\transposed \mathbf{b}_i\\
			&= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i
		\end{aligned}
	\end{equation}
	Since $\Delta \mathbf{R}$ is defined as rotating a point by at most $n_r$ (definition of geodesic distance $d_{\angle}(\mathbf{R}, \mathbf{R_c}) \leq n_r \iff d_{\angle}(\mathbf{I}, \Delta \mathbf{R} \leq n_r$) from this follows:
	\begin{equation}
		\begin{aligned}
			r_{min}^i &= \min_{\Delta \mathbf{R} \in \SOthree, \, d_{\angle}(\mathbf{I}, \Delta \mathbf{R}) \leq n_r} \left( ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i \right)\\ 
			&= ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\max(\theta - n_r, 0)) \\
			r_{max}^i &=
			\max_{\Delta \mathbf{R} \in \SOthree, \, d_{\angle}(\mathbf{I}, \Delta \mathbf{R}) \leq n_r} \left( ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 (\mathbf{R}_c^\transposed\mathbf{a}_i )^\transposed \Delta\mathbf{R}\mathbf{b}_i \right)\\ 
			&=  ||\mathbf{a}_i||^2 + ||\mathbf{b}_i||^2 - 2 ||\mathbf{a}_i|| \, ||\mathbf{b}_i|| \cos (\min(\theta + n_r, \pi))
		\end{aligned}
	\end{equation}

\begin{theorem}

	For the truncated least-squares function:

	\begin{equation}
		\label{eq:tls-t-opt-branch2}
		\begin{aligned}
			\text{TLS}(\mathbf{x}) = \sum_{i=1}^{N} \min \left(r(\mathbf{x}, \mathbf{x}_i)^2, \epsilon^2 \right)\\
		\end{aligned}
	\end{equation}

	where $\mathbf{x}_i$ are data points and $r_i = r(\cdot)^2$ are residual terms convex in $\mathbf{x}$. Given are the minimum and maximum value of each residual term $r_i$ over every $\mathbf{x} \in \mathcal{X}$, denoted as $r_{min}^i$ and $r_{max}^i$ respectively. Then, $\text{WLS}_{r}$ is a convex relaxation of \ref{eq:tls-t-opt-branch2} over $\mathcal{X}$:
	\begin{equation}
		\label{eq:tls-wls-relaxation}
		\begin{aligned}	
			\text{WLS}_{r}(\mathbf{x}) &= 
			\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right) + \sum_{i=1}^{N} o_i\epsilon^2\\
		\end{aligned}
	\end{equation}

	with the weights $w_i$ defined as:
	\begin{equation}
		\begin{aligned}	
			w_i  = 
			\begin{cases}
				\frac{\epsilon^2 - r_{min}^i}{r_{max}^i - r_{min}^i} &  r_{max}^i > \epsilon^2\\
				1 & \, \text{otherwise}
			\end{cases}
		\end{aligned}
	\end{equation}

	and the binary variables $o_i$ deciding whether the $i$-th residual is an outlier or not:
	\begin{equation}
		\begin{aligned}	
			o_i &= \begin{cases}
				1 &  r_{min}^i > \epsilon^2\\
				0 & \, \text{otherwise}
			\end{cases}
		\end{aligned}
	\end{equation}

\end{theorem}

\begin{proof}

	In the truncated-least-squares function, non-convexity is introduced solely by the truncation since the squared terms are convex.
	Therefore, we only need to show that the re-weighted sum leads to the residuals not being truncated. 
	For this, we apply an affine transform on each residual so that $w_i r_i + y_0 = r_{min}^i$ for $r_i = r_{min}^i$ and $w_i r_i + y_0 = \epsilon^2$ for $r_i = r_{max}^i$:
	Subtracting the first from the second equation yields:
	\begin{equation}
		\begin{aligned}	
			&w_i (r_{max}^i - r_{min}^i) = \epsilon^2 - r_{min}^i\\
			\iff &w_i  = \frac{\epsilon^2 - r_{min}^i}{r_{max}^i - r_{min}^i}
		\end{aligned}
	\end{equation}
	and solving for $y_0$ yields:
	\begin{equation}
		\begin{aligned}	
			w_i r_{min}^i + y_0 &= r_{min}^i\\
			\iff y_0 &= r_{min}^i - w_i r_{min}^i\\
			\iff y_0 &= (1- w_i) r_{min}^i
		\end{aligned}
	\end{equation}

	Therefore, $w_k r_k + (1 - w_k) r_{min}^k \leq \epsilon^2$ holds for all $k$ for which $o_k = 0$ and therefore these terms are not truncated and thus convex. The sum

	\begin{equation}
		\begin{aligned}	
			\sum_{i=1}^{N} (1 - o_i) \left(w_i r_i  + (1 - w_i) r_{min}^i \right)
		\end{aligned}
	\end{equation}

	is therefore convex since sums of convex functions are convex \cite[Ch. 2.3.2, p.36]{Boyd_Vandenberghe_2004}.

	Since $\sum_{i=1}^{N} o_i\epsilon^2$ is constant, the overall term \ref{eq:tls-wls-relaxation} also remains convex. Since $r_k \in [r^k_{min}, r^k_{max}]$ and $w_k \leq 1$, $w_k r_k  + (1 - w_k) r_{min}^k \leq r_k$ for all $k$ for which $o_k = 0$  Eq. \ref{eq:tls-wls-relaxation} is below the objective \ref{eq:tls-t-opt-branch2}. Therefore, \ref{eq:tls-wls-relaxation} is a convex relaxation of \ref{eq:tls-t-opt-branch2} over $\mathcal{X}$.

\section{Least-squares rotation estimation with equality-constrained angle}
\label{proof:wls-relax-eq-constrained-rotation}

\begin{theorem}

	The least-squares rotation estimation problem 
	\begin{equation}
		\label{eq:tls-rwls-a}
		\begin{aligned}
			\argmin_{\mathbf{R} \in \SOthree}  \quad &\sum_{i=1}^{N} w_i ||\mathbf{R}_c^\transposed \mathbf{a}_i - \Delta\mathbf{R} \mathbf{b}_i ||_2^2 \\
			\text{subject to} \quad &d_{\angle}(\mathbf{I}, \Delta\mathbf{R}) = n_r
		\end{aligned}
	\end{equation}
	where the angle of rotation is constrained to $n_r$ is equivalent to the following optimization problem over the rotation axis $\mathbf{n}$:
	\begin{equation}
		\label{eq:rot-est-constrained-qcqp}
		\begin{aligned}
			\argmin_{\mathbf{n} \in \Rthree} \quad &-\left(\mathbf{n}^\transposed \mathbf{A} \mathbf{n} + 2 \mathbf{g}^\transposed \mathbf{n} \right)\\
			\text{subject to} \quad &|| \mathbf{n}|| = 1\\
		\end{aligned}
	\end{equation}
	with
	\begin{equation}
		\label{eq:ev-constr-data-mat-a}
		\begin{aligned}
			\mathbf{B} &= \sum_{i=1}^N w_i \mathbf{a}_i\mathbf{b}_i^\transposed \in \Rone^{3 \times 3}, \quad
			\mathbf{z} = \sum_{i=1}^N w_i [\mathbf{a}_i]_{\times}\mathbf{b}_i \in \Rone^{3}, \quad \mathbf{C} = \mathbf{B} + \mathbf{B}^\transposed\\
			\mathbf{A} &= \sin^2 \left(\frac{n_r}{2} \right) \mathbf{C}, \quad \mathbf{g} =  \sin \left(\frac{n_r}{2} \right) \cos \left(\frac{n_r}{2} \right) \mathbf{z}
		\end{aligned}
	\end{equation}
\end{theorem}

\begin{proof}

	We first separate the optimization over the rotation axis and angle. For this, we use the quaternion representation of \ref{eq:tls-rwls-a}. One can show that the problem \ref{eq:tls-rwls-a} is equivalent to:

	\begin{equation}
		\label{eq:davenports-q-problem}
		\begin{aligned}
			\max_{\mathbf{q} \in \Rone^4} &\quad \mathbf{q}^\transposed \mathbf{K} \mathbf{q}\\
			\text{subject to} &\quad  ||\mathbf{q}|| = 1
		\end{aligned}
	\end{equation}

	where $\mathbf{q}$ is a 4-dimensional vector representing the quaternion of rotation and $\mathbf{K} \in \Rone^{4\times4}$ is a data matrix. This optimization problem is known as the \textit{Davenport's q-method}. The quaternion can be written in axis-angle form as:

	\begin{equation}
		\label{eq:q-axis-angle}
		\begin{aligned}
			\mathbf{q} = \left( \sin \left( \frac{\theta}{2} \right) \mathbf{n}, \cos \left(\frac{\theta}{2} \right) \right)^\transposed
		\end{aligned}
	\end{equation}
	i.e. the first three elements are the unit-vector $\mathbf{n} \in \Rthree, ||\mathbf{n}|| = 1$ of the axis of rotation multiplied by $\sin \left( \frac{\theta}{2} \right)$, while the fourth element is just $\cos \left(\frac{\theta}{2} \right)$ and $\theta$ is the angle of rotation.
	The data matrix $\mathbf{K}$ has the following block-structure:

	\begin{equation}
		\label{eq:k-mat}
		\begin{aligned}
			\mathbf{K} = \left(
			\begin{array}{c|c}
				\begin{array}{ccc}
					& & \\
					& \mathbf{C} & \\
					& & 
				\end{array}
				&
				\begin{array}{c}
					\mathbf{z} \\
				\end{array}
				\\
				\hline
				\begin{array}{ccc}
					& \mathbf{z}^\transposed &
				\end{array}
				&
				\trace(\mathbf{C})
			\end{array}
			\right) \in \Rone^{4 \times 4}
		\end{aligned}
	\end{equation}

	We can rewrite the objective function $\mathbf{q}^\transposed \mathbf{K} \mathbf{q}$  of \ref{eq:davenports-q-problem} using the blocks of the block-matrix $\mathbf{K}$:

	\begin{equation}
		\label{eq:dv-q-rew}
		\begin{aligned}
			&\mathbf{q}^\transposed \mathbf{K} \mathbf{q}\\
			&= \sin^2 \left(\frac{\theta}{2} \right) \mathbf{n}^\transposed \mathbf{C} \mathbf{n} + 2 \sin \left(\frac{\theta}{2} \right) \cos \left(\frac{\theta}{2} \right) \mathbf{z}^\transposed \mathbf{n} + \cos^2 \left(\frac{\theta}{2} \right) \trace(\mathbf{C})
		\end{aligned}
	\end{equation}

	We now have a scalar term where the axis $\mathbf{n}$ and the angle $\theta$ occur separately and with this we can formulate the optimization problem over the axis. By setting $\theta = n_r$ and solving only over $\mathbf{n}$ we can therefore solve the equality-constrained problem \ref{eq:tls-rwls-a}.

	Since the term $ \cos^2 \left(\frac{\theta}{2} \right) \trace(\mathbf{C})$ is constant in Eq. \ref{eq:dv-q-rew}, it does no affect the minimizer and we will leave it out. 
	By defining the data-matrices as in \ref{eq:ev-constr-data-mat-a}, we immediatelly see that the problem 
	\ref{eq:rot-est-constrained-qcqp} is equivalent to Davenports Q-problem \ref{eq:davenports-q-problem} and therefore equivalent to \ref{eq:tls-rwls-a}, concluding the proof.
\end{proof}
